Rheology of Cubic Blue Phases

### 2013-07-02 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2034 40876441-3496-0410-a613-a2fca35b2ae2

\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}
\pagestyle{fancy}

@@

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up 
\rev{
into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction according to the imposed flow.
}
BPI leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- very slow flow leads to a transition into an amorphous network 
with an apparent yield stress. Larger shear rates can make the network sensitive
to flow-induced transitions to metastable and amorphous states. For even
larger flow rates a Grandjean texture is observed, just as in the case of BPII.
At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.
Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
its helical axis, small pressure differences drive flow mainly through
``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
\rev{
Early experiments with cholesteric liquid crystals
showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
}
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
\rev{
These result, however, were derived under the 
assumption that the molecules rotate only in flow-gradient plane whilst
the orientation of the cholesteric helix remained unchanged.
}  
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

@@

The system of coupled partial differential equations~\ref{op-eom}
and~\ref{NSE} is solved by means of a hybrid method~\cite{Marenduzzo:2007}
which uses a combination of lattice Boltzmann and finite difference schemes.
(This is in contrast with some earlier methods using solely
LB~\cite{Denniston:2001, Denniston:2004}.)
The Navier-Stokes
equation is solved via the lattice Boltzmann approach, using a standard
three-dimensional model with 19 discrete velocities (D3Q19).
A regular lattice with spacing $\Delta x = \Delta y = \Delta z = 1$ is
used and the time step is $\Delta t = 1$ in lattice units.
Coupling to the thermodynamic sector is via a
local body force computed as the divergence of the thermodynamic
stress Eq.~\ref{Pi}; the resulting velocity field is used in the computation
of the time evolution of $\mathbf{Q}$ via a standard finite difference
method using the same grid and the same time step as the LB. The system
is periodic in all three coordinate directions. Sliding
boundary conditions, or Lees-Edwards planes, are explicitly implemented
in both hydrodynamic and thermodynamic sectors to impose shear on the
system. On
crossing a Lees-Edwards boundary a Galilean transformation is applied
with a velocity increment which is fixed in time (and limited in size by the
low Mach number constraint of LB). An appropriate transformation
of LB distributions which propagate across a boundary~\cite{Wagner:2002}
must be applied at each time step, and appropriate
adjustment to $W_{\alpha\beta}$ is required to compute cross-plane
gradients of the velocity field used in the update to ${\mathbf Q}$.
In both cases interpolation of the
relevant quantities is required to cope with the relative displacement
of neighbouring lattice sites separated by a sliding plane (the
displacement may be a fraction of a lattice unit at any given time
step). The use of multiple sliding planes equally spaced in a single
system allows
the overall shear rate to be maintained indefinitely as the system
becomes larger
in the velocity gradient direction (in contrast with the use of solid
walls to impose shear). This hybrid method with Lees-Edwards planes
has been used successfully to study e.g., smectics~\cite{Henrich:2012a}.

In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
phases BPI and BPII.
Typical runtimes for system size $L_x\times L_y\times L_z=128^3$ on 512 processes were in the region of 18 to 24 hours.  
The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found elsewhere~\cite{Henrich:2011a,Henrich:2010b}.
In what follows, $x$, $y$ and $z$ denote respectively the velocity, velocity
gradient and vorticity direction; $\Pi_{xy}$ is therefore the shear stress.

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
\rev{
For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) that
may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
GJ and FAN occur regardless of the initial state.
}

@@

As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
\caption{
\rev{
Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
of BPII versus strain $\gamma = \gd\cdot t$  
normalised to the background viscosity $\eta$.
The pictures show the regimes BPII-1 (top, PRC), BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.
}
} 
\label{bp2-appvisc}
\end{figure}

Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
\rev{For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
the data to a very good approximation.
This holds true even in regime BPII-2, where the network breaks up and 
the liquid crystal adopts a Grandjean structure. The coefficients of the power law fit are
here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
remarkably small in BPII, in the range of shear rates which we have explored here.}

\rev{
Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
over total strain.
A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
without any additional contribution from the liquid crystal.
The top picture shows data for regime BPII-1, where the periodic breakup and reconnecting of the network 
in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
absence at very low shear rate of a permeation mode is probably due to our
choice of boundary conditions.
}

%{We note that permeation flows, which would
%{\em not} lead to oscillatory response, are effectively disallowed by the combination
%of the geometry and boundary condition we consider. This is
%because the periodic boundary conditions impose macroscopic distortion
%of the network, and are effective analogous to infinitely distant
%but anchored wall along the $xy$ plane.
%This means that  our results are effectively valid for sheared bulk 
%BPs which are anchored at the walls. Furthermore, permeation flows are 
%only restricted to very slow flows, and are unstable for intermediate 
%and high flows, where the response of the network depends less on
%the details of the boundary conditions used at 
%the wall~\cite{Marenduzzo:2006a,Marenduzzo:2006b}.}

\rev{
We observe another configuration, lying between those with periodically recurring conformations 
and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.
}

\rev{
The picture at the bottom of Fig.~\ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
the flow-aligned nematic state. We will address these regimes in section \ref{gj-fan}, as they are common to BPI and BPII.
}

\rev{
The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
to the tumbling motion of the local director field, which is absent in the Grandjean texture.
For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
}

@@

@@

Interestingly, while being displaced with the flow the entire network moves 
along the vorticity direction (z) as well.  A similar behaviour has been recently 
observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
In contrast to the stick-slip motion that has been reported there,
in this case the movement is steady. 
The vorticity motion occurs in such a way that the positions of breakup and reconnection 
point in the network, visible in Fig. \ref{bp2-1-disc}, are slightly offset and allow the network 
to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
is $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
for the network to move one unit cell along $z$ (the vorticity direction).

@@

\begin{figure*}[htpb]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902r.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903r.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
The view is along the x-direction. This means the velocity ${\mathbf v}$ 
has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with right-handed helicity ($q_0>0$ in Eq.~\ref{FE}) travels rightwards, whereas the one one the right,
with reversed helicity, moves leftward. 
The periodicity of the motion along the vorticity direction is six times longer than the time 
it takes the network to reconnect along the flow direction.}
\label{bp2-1-velo}
\end{figure*}

@@

Table~\ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. 
\rev{
Therefore the averaged flow along the vorticity
direction is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative. This claim is further supported by the characteristic 
velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
whereas the network as a whole moves only in one direction.
}

@@

\rev{
Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
shear rates as in Fig.~\ref{bp2-flowcurve}, whereas Fig. \ref{bp1-appvisc} shows 
the apparent viscosity ration $\eta_{app}/\eta$ over total strain.
}

For intermediate (but not low) shear rates
($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more find regular oscillations in the stress
response versus time, and in the network morphology -- we label this 
regime BPI-2. 
\rev{
This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.
}

The BPI-2 regime now does not persist to arbitrarily low shear: there the
oscillations become irregular and seemingly chaotic, and after a few
cycles the disclination pattern becomes amorphous.
\rev{
For these very low shear rates 
($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
the flow curve in Fig. \ref{bp1-flowcurve}
apparently develops a horizontal yield stress branch. 
We refer to this flow regime as BPI-1.
}

The BPI-2 regime is also unstable if the shear is increased {\em above} 
a critical value 
\rev{
in the region of Ericksen numbers $Er \simeq O(1)$.
This is a regime 
($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
where viscous and elastic forces have more or less 
the same magnitude we labelled BPI-3. 
In this window of strain rates the network seems to become
increasingly sensitive towards small perturbations in the flow field
that can cause irregularities in the order parameter structure.
Due to the back flow, i.e. the coupling of flow and order in our model, 
these distortions form a possible route towards an instance of 
"rheochaos"~\cite{rheochaos,Cates:2002}.
}

\rev{
Finally, for shear rates such that
$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
and before breakup into a flow-aligned nematic occurs 
(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
we observe the same cholesteric Grandjean configuration with the 
helical axis along the gradient direction as 
reported above for BPII-2 (see
section \ref{gj-fan}). This regime is referred to as BPI-4.
}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
\rev{
Depending on the configuration in steady shear flow five different regimes 
can be identified: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
steady flow with periodically recurring patterns (BPI-2, open squares), 
breakup of the disclination network into amorphous state (BPI-3, open circles)
\com{Calculate another shear rate to get a regime},
cholesteric helix with axis in gradient direction, also referred to
as Grandjean texture and frustrated metastable states that transform into 
a Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
and flow-aligned nematic state (BPI-5, solid triangles). 
The error bars represent the minimum and maximum shear stress 
during one cycle in regime BPI-2.
The two dashed lines show fits to the data in regime BPI-2 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
and in regime BPI-4 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
}
\label{bp1-flowcurve}
\end{figure}

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_a.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_b.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_c.pdf}
\caption{
\rev{
Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised by the background viscosity $\eta$: 
The top picture shows regime BPI-1 (AN) where
the BP network transforms into an amorphous network featuring a yield stress.
The centre picture shows regimes BPI-2 (PRC) and BPI-3 (AN), 
whereas the picture at the bottom
depicts data of the Grandjean texture (GJ and FGJ if frustrated) in regime BPI-4 
and the flow-aligned nematic state (BPI-5, FAN). 
Note that the $\eta_{app}/\eta$ is negative for the lowest 
flow rates as the (negative) contribution of the elastic forces
of the network dominates the total stress.
}
}
\label{bp1-appvisc}
\end{figure}

@@

In this Appendix we give some more details on our simulations,
including the relationship between boundary conditions and
permeation flow.

>>>>

\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}

\pagestyle{fancy}

@@

\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
For larger shear rates, the structure breaks up 
\rev{
into a Grandjean texture with a 
cholesteric helix lying along the flow gradient direction with respect to the imposed flow.
}
Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
shear rates -- very slow flow leads to a transition into an amorphous network 
with an apparent yield stress. 
\rev{
Larger shear rates can make the network sensitive
to flow-induced transitions to metastable and amorphous states.
}
For even larger flow rates the same break-up into a Grandjean texture as for Blue Phase II is observed.
At the highest imposed flow rates both cubic blue phases adopt a flow-aligned nematic state.
Our results provide the first theoretical investigation of sheared blue phases in large systems, 
and are relevant to understanding the bulk rheology of these materials.}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

@@

Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
its helical axis, small pressure differences drive flow mainly through
``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
In the permeation mode the liquid crystal flows while leaving the director
field virtually unchanged, which leads to high dissipation and large
viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
They also found that a strong secondary flow appears.
\rev{
Early experiments with cholesteric liquid crystals
showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
}
Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
\rev{
However, this result was derived under the 
assumption that the molecules rotate only in flow-gradient plane whilst
the orientation of the cholesteric helix remained unchanged.
}  
Rey also studied cholesterics subjected to both steady flow and low frequency
small amplitude oscillatory shear for different helix orientations
~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
excited when the helix was aligned along the flow direction; splay/bend
deformation occurred when the helix was aligned along the velocity gradient;
but only twist deformations
appeared when the helix was aligned along the vorticity axis.

@@

The system of coupled partial differential equations~\ref{op-eom}
and~\ref{NSE} is solved by means of a hybrid method~\cite{Marenduzzo:2007}
which uses a combination of lattice Boltzmann and finite difference schemes.
(This is in contrast with some earlier methods using solely
LB~\cite{Denniston:2001, Denniston:2004}.)
The Navier-Stokes
equation is solved via the lattice Boltzmann approach, using a standard
three-dimensional model with 19 discrete velocities (D3Q19).
A regular lattice with spacing $\Delta x = \Delta y = \Delta z = 1$ is
used and the time step is $\Delta t = 1$ in lattice units.
Coupling to the thermodynamic sector is via a
local body force computed as the divergence of the thermodynamic
stress Eq.~\ref{Pi}; the resulting velocity field is used in the computation
of the time evolution of $\mathbf{Q}$ via a standard finite difference
method using the same grid and the same time step as the LB. The system
is periodic in all three coordinate directions. Sliding
boundary conditions, or Lees-Edwards planes ~\cite{Wagner:2002}, are explicitly implemented
in both hydrodynamic and thermodynamic sectors to impose shear on the
system. 
\rev{
The same hybrid method with Lees-Edwards planes
has been used successfully to study binary mixtures, particle suspensions~\cite{MacMeccan:2009, Aidun:2010}, 
emulsions~\cite{Frijters:2012} and smectic liquid crystals~\cite{Henrich:2012a}.
A more detailed discussion of our use of Lees-Edwards boundary conditions can be found in the Appendix 
of this paper.
}

In the following we report results of simulations of the bulk flow behaviour of the cubic Blue 
Phases BPI and BPII.
Typical runtimes for system size $L_x\times L_y\times L_z=128^3$ on 512 processes were in the region of 18 to 24 hours.  
The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
is equal to about $10^8$~Pa. Further details about the conversion 
from LBU to SI units can be found elsewhere~\cite{Henrich:2011a,Henrich:2010b}.
In what follows, $x$, $y$ and $z$ denote respectively the velocity, velocity
gradient and vorticity direction; $\Pi_{xy}$ is therefore the shear stress.

@@

Simple shear flow was imposed by means of the Lees-Edwards boundary
conditions with
the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
velocity gradient along the $y$-direction.
The shear rates were varied over more than three orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $3.75\times\e{-3}$ LBU.
\rev{
For clarity we classify various flow regimes, namely three in the case of Blue Phase II and five in the 
case of Blue Phase I. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) which 
may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
GJ and FAN occur in both blue phases as steady states regardless of the initial state.
}

@@

As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
Typical choices are $q_s=0.17$ for BPI and $q_s=0.15$ for BPII.
(Note that $q_s$ is small but non-zero at the disclination core. 
The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
\caption{
\rev{
Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
of BPII versus strain $\gamma = \gd\cdot t$  
normalised to the background viscosity $\eta$.
The pictures show the regimes BPII-1 (periodically recurring conformations, top), BPII-2 (Grandjean texture, bottom) and BPII-3 (flow-aligned nematic, bottom). 
The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without any additional contribution of the liquid crystal, i.e. total shear-thinning.
}
} 
\label{bp2-appvisc}
\end{figure}

Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
\rev{
Time averages were taken as average over one stress cycle in the steady state.
Where no oscillations occurred we used the final value at the end of the run.
For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
the data to a very good approximation.
This holds true even in regime BPII-2, where the network breaks up and 
the liquid crystal adopts a Grandjean structure with the helical axis oriented along the
flow gradient direction ($y$). The coefficients of the power law fit are
here $a=0.68$ and $b=0.94$. Hence, for the range of shear rates which we have explored here, 
the degree of shear-thinning is remarkably small in BPII.
}

\rev{
Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
over total strain.
A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
without any additional stress contribution from the liquid crystal.
The top picture shows data for regime BPII-1, where the periodic break-up and reconnecting of the network 
in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
absence of a permeation mode at very low shear rates may be due to our
choice of boundary conditions. We refer the reader to the Appendix for a detailed discussion.
}

\rev{
We observe another configuration, lying between those with periodically recurring conformations 
and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
and the liquid crystal is flowing in ``nematic planes'' with the director predominantly oriented in flow-vorticity plane and having only
a small angle with the flow direction.
The picture at the bottom of Fig.~\ref{bp2-appvisc} shows data for this regime, to which we refer as BPII-2. The 
curves for the two largest flow rates are for BPII-3, the flow-aligned nematic state. 
We will address these regimes in section \ref{gj-fan}, as they are common to both cubic blue phases BPI and BPII.
}

\rev{
The travelling helical waves which was identified in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
the vorticity axis turns out to be a metastable flow state as it features larger free energies, higher stresses and dissipation due 
to the tumbling motion of the director field, which is completely absent in the Grandjean texture.
For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
}

@@

\begin{figure}[htbp]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902r.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903r.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
The view is along the x-direction. This means the velocity ${\mathbf v}$ 
has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with right-handed helicity ($q_0>0$ in Eq.~\ref{FE}) travels rightwards, whereas the one one the right,
with reversed helicity, moves leftward. 
The periodicity of the motion along the vorticity direction is six times longer than the time 
it takes the network to reconnect along the flow direction.}
\label{bp2-1-velo}
\end{figure}

@@

Interestingly, while being displaced with the flow the entire network moves 
along the vorticity direction ($z$) as well.  A similar behaviour has been recently 
observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
In contrast to the stick-slip motion that has been reported there,
in this case the movement is steady. 
The vorticity motion occurs in such a way that the positions of breakup and reconnection 
point in the network, visible in Fig. \ref{bp2-1-disc}, are slightly offset and allow the network 
to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
is $\tau_V=6\tau_F$, 
i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
for the network to move one unit cell along $z$ (the vorticity direction).

@@

@@

Table~\ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. 
\rev{
Therefore the averaged flow along the vorticity
direction is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative. This claim is further supported by the characteristic 
velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
The secondary flow in vorticity direction takes clearly place in positive {\em and} 
negative z-direction, whereas the movement of the network as a whole is only in one direction.
}

@@

\rev{
Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
shear rates as in Fig.~\ref{bp2-flowcurve} and Fig. \ref{bp1-appvisc} depicts 
the apparent viscosity ratio $\eta_{app}/\eta$ over total strain.
}

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
\rev{
Depending on the configuration in steady shear flow five different regimes 
can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
steady flow with periodically recurring patterns (BPI-2, open squares), 
break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles),
cholesteric helix with axis in gradient direction, also referred to
as Grandjean texture as well as frustrated metastable Grandjean-like states that can transform into 
a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
and flow-aligned nematic state (BPI-5, solid triangles). 
The error bars represent the minimum and maximum shear stress 
during one cycle in regime BPI-2.
The two dashed lines show fits to the data in regime BPI-2 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
and in regime BPI-4 with 
$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
}
\label{bp1-flowcurve}
\end{figure}

For intermediate (but not too low) shear rates
($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more 
find regular oscillations in the network morphology and in the stress
response versus time. We label this regime BPI-2. 
\rev{
The centre picture of Fig.~\ref{bp1-appvisc} shows the corresponding data for the
apparent viscosity ratio versus strain.
}

The BPI-2 regime now does not persist to arbitrarily low shear: there the
oscillations become irregular and seemingly chaotic, and after a few
cycles the disclination pattern becomes amorphous.
\rev{
For these low shear rates 
($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
the flow curve in Fig. \ref{bp1-flowcurve}
develops a horizontal yield stress branch. 
We refer to this flow regime as BPI-1.
}

The BPI-2 regime is also unstable if the shear is increased {\em above} 
a critical value, 
\rev{
which is in the region of Ericksen numbers $Er \simeq O(1)$.
This is a regime 
($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
we labelled BPI-3. It is characterised by the balance between viscous and elastic forces.
This has an influence on the stability of the disclination network as it 
seems to become increasingly sensitive to small perturbations in the flow field
which couple back to the order parameter dynamics and vice versa. This phenomenon forms 
a possible route towards an instance of ``rheochaos'' ~\cite{rheochaos,Cates:2002}.
}

\rev{
Finally, for shear rates such that
$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
and before the break-up into a flow-aligned nematic occurs 
(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
we observe the same cholesteric Grandjean configuration with the 
helical axis along the gradient direction as 
reported above for BPII-2 (see
section \ref{gj-fan}). This regime is referred to as BPI-4.
}

\begin{figure}[htpb]
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_a.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_b.pdf}\\
\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_c.pdf}
\caption{
\rev{
Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised to the background viscosity $\eta$: 
The top picture shows regime BPI-1 (amorphous network) where
the BP network transforms into an amorphous network featuring a yield stress.
The centre picture shows regimes BPI-2 (periodically recurring conformations) and BPI-3 (amorphous network), 
whereas the picture at the bottom
depicts data of the (partly frustrated) Grandjean texture in regime BPI-4 
and the flow-aligned nematic state, labelled BPI-5 in this context. 
Note that the $\eta_{app}/\eta$ is negative for the lowest 
flow rates as the (negative) contribution of the elastic forces
of the network dominates the total stress.
}
}
\label{bp1-appvisc}
\end{figure}

@@
